Human capital mediates natural selection in contemporary humans

### 2022-05-04 why-natural-selection.Rmd
Really use medians for percentage differences

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
          !! score_col := pretty_names(!! score_col),
          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
        ) %>% 
        rename(p = p.value) %>% 
        ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
          my_vline + 
          confint_geom_segment +
          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
          pval_scale() +
          my_fill_scale(n = n_cats, direction = fill_direction) +
          my_labs() +
          my_theme() +
          scale_linetype_manual(name = "", 
                                  values = c("95% c.i. uncorrected" = 1)) +
          guides(
            fill     = guide_legend(order = 1),
            stroke   = guide_legend(order = 2),
            linetype = guide_legend(order = 3)
          )
}

@@

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
                    color = "grey60", alpha = 0.25, size = 1.1, 
                    lineend = "round")
}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey60", 
                       size = 3) +
      geom_point(aes(y = score), size = 1.5) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c(
                                      "-" = "triangle down open", 
                                      "+" = "triangle open", 
                                      "o" = "circle open"
                                    )) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

mean_pct_diff <- function (dfr) {
   prop <- dfr %>% filter(
                    p.value.x < 0.05/33,
                    sign(estimate.x) == sign(estimate.y)
                  ) %>% 
                  mutate(
                    pct_diff = estimate.x/estimate.y
                  ) %>% 
                  pull(pct_diff) %>% 
                  mean()
   # percentage difference:
   prop - 1
}

pct_diff_edu <- mean_pct_diff(res_edu_wide)
pct_diff_income <- mean_pct_diff(res_income_wide)
pct_diff_n_partners <- mean_pct_diff(res_partners_wide)
pct_diff_with_partner <- mean_pct_diff(res_with_partner_wide)
```

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
education category, and for the lowest income category. The mean percentage
difference between the lowest and highest education categories, among scores
which are significant for the lowest category and have the same sign across
categories, is `r scales::percent(pct_diff_edu)`. Between the lowest and
highest income categories, it is `r scales::percent(pct_diff_income)`. These
results are robust to controlling for respondents' age (Appendix section
\@ref(sec:age-control)). Turning to family structure, we split respondents by
lifetime number of sexual partners, at the median value of 3 (Figure
\@ref(fig:plot-n-partners)a). Now, selection effects are larger and more
significant among those with more than 3 lifetime partners, with a mean percentage
difference of `r scales::percent(pct_diff_n_partners)`. Next we split
respondents by whether they were living with a spouse or partner at the time of
interview (Figure \@ref(fig:plot-n-partners)b).  Effects are larger among those not living with a spouse or partner. The mean percentage difference is 
`r scales::percent(pct_diff_with_partner)`.[^sex]

@@

Polygenic scores which correlate with high (low) earnings and more (less)
education are being selected against (selected for). In addition, many of the
phenotypes under positive selection are linked to disease risk. Many people
would probably prefer to have high educational attainment, a low risk of ADHD
and major depressive disorder, and a low risk of coronary artery disease, but
natural selection is pushing against genes associated with these traits.
Potentially, this could increase the health burden on modern populations, but
that depends on effect sizes. Our results show that naïve estimates can be
affected by sample ascertainment bias. There may be remaining sources of
ascertainment bias after our weighting; if so, we expect that, like the sources
of ascertainment we have controlled for, they probably bias our results towards
zero. Researchers should be aware of the risks of ascertainment when studying 
modern natural selection.

@@

res_sex_wide <- res_sex %>% 
      select(term, estimate, sex, diff.p.value) %>% 
      pivot_wider(names_from = "sex", values_from = "estimate") %>% 
      mutate(
        min_est = pmin(Female, Male),
        max_est = pmax(Female, Male),
        term = pretty_names(term),
        term = fct_reorder(term, Female)
      ) 
n_comps <-as.double(nrow(res_sex_wide))

res_sex %>% 
      mutate(
        term = pretty_names(term),
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p < 0.05/{{n_regs}})) +
        my_vline +
        geom_linerange(data = res_sex_wide, 
          aes(
            y      = term,
            x      = NULL,
            fill   = NULL,
            xmin   = min_est, 
            xmax   = max_est, 
            colour = diff.p.value < 0.05/{{n_comps}}
          ), 
        group = "") +
        pval_geom_point() +
        pval_scale(name = "p < 0.05 (corrected)") +
        my_fill_scale(n = 2) +
        my_labs() +
        my_theme()

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = `N partners`, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
        sexes_guides

@@

res_with_partner_sex %>% 
      mutate(
        score_name = pretty_names(score_name),
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = Household, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(
                                name   = "", 
                                values = c("95% c.i. uncorrected" = 1)
                              ) +
        sexes_guides

@@

standard_ggplot(res_fe_fertility, fill_col = Regression, n_regs = n_regs,
                  fill_direction = -1)

@@

With a reduced sample size, all within-sibling effects are insignificant after
Bonferroni correction. However, effect sizes are positively correlated with
effect sizes from the pooled model, and about 50% smaller (regressing
within-sibling on pooled effect sizes, $b$ = 
`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the
~50% decrease in heritability in
within-sibling GWASs on age at first birth and educational attainment
[@Howe2021.03.05.433935]. We see these results as providing tentative evidence
that polygenic scores cause fertility, with effects being partly driven by
correlations with environmental variation in human capital. We also reran
within-siblings regressions adding a control for education. Most effect sizes
barely change, suggesting that our measure of education does not in general
mediate differences in fertility among siblings.

@@

```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores. Parental generation weights multiplied by 1/number of siblings.", fig.height = 6}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term     = pretty_names(term),
        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var  = fct_inorder(dep.var),
        dep.var  = fct_recode(dep.var, 
                                `Respondents' RLRS` = "RLRS",
                                `Parents' RLRS` = "RLRS_parents" 
                              ),
        reg.type = fct_recode(reg.type, 
                               `Resid. for 100 PCs`  = "controlled", 
                               Unresidualized        = "raw"
                             )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline + 
        facet_grid(rows = vars(dep.var), scales = "free") +
        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered() +
        theme(
          axis.text.x  = element_text(size = 7),
          legend.text  = element_text(size = 8),
          legend.title = element_text(size = 9)
        )

@@

res_pcs %>% 
      mutate(
        estimate = abs(estimate),
        dep.var  = fct_relevel(dep.var, "RLRS"),
        dep.var  = fct_recode(dep.var,
          "Parents"     = "RLRS_parents",
          "Respondents" = "RLRS"
        ),
        top_pcs  = ifelse(as.numeric(pc_name) <= 5, term, "")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(estimate, dep.var, colour = p < 0.05/{{n_regs}})) + 
        geom_point(alpha = 0.8, position = pj) +
        geom_text(aes(estimate, dep.var, label = top_pcs), position = pj, 
                    hjust = "right", color = "black", size = 3) +
        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
        labs(x = "Effect size (log scale)", y = "") + 
        pval_scale(aesthetics = "colour")

>>>>

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
    scale_shape_manual(values = 1:n_cats + 14) +
    my_labs() +
    my_theme() +
    confint_scale +
    scale_alpha_discrete(range = c(1, 0), 
                           guide = guide_legend(
                             order = 0,
                             override.aes = list(
                               shape = c(1, 16),
                               color = "black",
                               alpha = 1,
                               size = 1.8
                             )
                         )) 
}

@@

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
                    color = "grey55", alpha = 0.2, size = 1.1, 
                    lineend = "round")
}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey70", 
                       size = 3) +
      geom_point(aes(y = score), size = 1.5) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c(
                                      "-" = "triangle down open", 
                                      "+" = "triangle open", 
                                      "o" = "circle open"
                                    )) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

median_pct_diff <- function (dfr) {
   prop <- dfr %>% filter(
                    p.value.x < 0.05/33,
                    sign(estimate.x) == sign(estimate.y)
                  ) %>% 
                  mutate(
                    pct_diff = estimate.x/estimate.y
                  ) %>% 
                  pull(pct_diff) %>% 
                  median()
   # percentage difference:
   prop - 1
}

pct_diff_edu <- median_pct_diff(res_edu_wide)
pct_diff_income <- median_pct_diff(res_income_wide)
pct_diff_n_partners <- median_pct_diff(res_partners_wide)
pct_diff_with_partner <- median_pct_diff(res_with_partner_wide)
```

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
education category, and for the lowest income category. The median percentage
difference between the lowest and highest education categories, among scores
which are significant for the lowest category and have the same sign across
categories, is `r scales::percent(pct_diff_edu)`. Between the lowest and
highest income categories, it is `r scales::percent(pct_diff_income)`. These
results are robust to controlling for respondents' age (Appendix section
\@ref(sec:age-control)). Turning to family structure, we split respondents by
lifetime number of sexual partners, at the median value of 3 (Figure
\@ref(fig:plot-n-partners)a). Now, selection effects are larger and more
significant among those with more than 3 lifetime partners, with a median percentage
difference of `r scales::percent(pct_diff_n_partners)`. Next we split
respondents by whether they were living with a spouse or partner at the time of
interview (Figure \@ref(fig:plot-n-partners)b).  Effects are larger among those not living with a spouse or partner. The median percentage difference is 
`r scales::percent(pct_diff_with_partner)`.[^sex]

@@

Polygenic scores which correlate with lower earnings and less education are
being selected for. In addition, many of the phenotypes under positive selection
are linked to disease risk. Many people would probably prefer to have high
educational attainment, a low risk of ADHD and major depressive disorder, and a
low risk of coronary artery disease, but natural selection is pushing against
genes associated with these traits. Potentially, this could increase the health
burden on modern populations, but that depends on effect sizes. Our results show
that naïve estimates can be affected by sample ascertainment bias. There may be
remaining sources of ascertainment bias after our weighting; if so, we expect
that, like the sources of ascertainment we have controlled for, they probably
bias our results towards zero. Researchers should be aware of the risks of
ascertainment when studying modern natural selection.

@@

res_sex_wide <- res_sex %>% 
      select(term, estimate, sex, diff.p.value) %>% 
      pivot_wider(names_from = "sex", values_from = "estimate") %>% 
      mutate(
        min_est = pmin(Female, Male),
        max_est = pmax(Female, Male),
        term = pretty_names(term),
        term = fct_reorder(term, Female),
        p = diff.p.value
      ) 
n_comps <-as.double(nrow(res_sex_wide))

res_sex %>% 
      standard_ggplot(fill_col = sex, n_regs = n_regs, score_col = term,
                        conf_int = FALSE) +
      geom_linerange(data = res_sex_wide, inherit.aes = FALSE, 
                     color = "steelblue4", alpha = 1,
              aes(
                y      = term,
                x      = NULL,
                xmin   = min_est + 0.0005, 
                xmax   = max_est - 0.0005,
                linetype = p < 0.05/{{n_comps}}
              )) +
      scale_linetype_manual(values = c("TRUE" = 1, "FALSE" = 2))

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      standard_ggplot(fill_col = `N partners`, score_col = score_name, 
                        n_regs = n_regs) +
        facet_grid(cols = vars(sex))

@@

res_with_partner_sex %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      standard_ggplot(fill_col = Household, score_col = score_name, 
                        n_regs = n_regs) +
        facet_grid(cols = vars(sex))

@@

res_fe_fertility %>% 
  mutate(
    Controls = ifelse(Regression == "Controlled", "FE + education", "FE only")
  ) %>% 
  standard_ggplot(fill_col = Controls, n_regs = n_regs,
                  fill_direction = -1) +
    guides(alpha = guide_legend(
                   override.aes = list(
                             shape = 1,
                             color = "black",
                             alpha = 1,
                             size = 1.8
                   )
                 )
    )

@@

With a reduced sample size, all within-sibling effects are insignificant after
Bonferroni correction. However, effect sizes are positively correlated with
effect sizes from the pooled model, and about 70% smaller (regressing
within-sibling on pooled effect sizes, $b$ = `r format(coef_pool_w, digits = 3)`).
This attenuation is broadly consistent with the decrease in heritability
in within-sibling GWASs on age at first birth and educational attainment
[@Howe2021.03.05.433935]. We see these results as providing tentative evidence
that polygenic scores cause fertility, with effects being partly driven by
correlations with environmental variation in human capital. We also reran
within-siblings regressions adding a control for education. Most effect sizes
barely change, suggesting that our measure of education does not in general
mediate differences in fertility among siblings.

@@

```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores. Parental generation weights multiplied by 1/number of siblings.", fig.subcap = c("Respondents", "Parents"), fig.ncol = 1}

@@

n_regs <- as.double(nrow(res_all))
res_all %<>% mutate(
               `PGS` = ifelse(reg.type == "controlled", 
                                "Residualized for 100 PCs", "Unresidualized")
             )

res_all %>% 
      filter(dep.var == "RLRS") %>% 
      standard_ggplot(fill_col = PGS, n_regs = n_regs)

res_all %>% 
      filter(dep.var == "RLRS_parents") %>% 
      standard_ggplot(fill_col = PGS, n_regs = n_regs)

@@

res_pcs %>% 
      mutate(
        estimate = abs(estimate),
        dep.var  = fct_relevel(dep.var, "RLRS"),
        dep.var  = fct_recode(dep.var,
          "Parents"     = "RLRS_parents",
          "Respondents" = "RLRS"
        ),
        top_pcs  = ifelse(as.numeric(pc_name) <= 5, term, "")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(estimate, dep.var, shape = p < 0.05/{{n_regs}})) + 
        geom_point(alpha = 0.8, color = "steelblue4", position = pj) +
        geom_text(aes(estimate, dep.var, label = top_pcs), position = pj, 
                    hjust = "right", color = "black", size = 3) +
        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
        labs(x = "Effect size (log scale)", y = "") + 
        scale_shape_manual(values = c("FALSE" = "circle open", "TRUE" = "circle"))
